---
title: "Gathering_Data"
author: "Marcela Aragon"
date: "2026-07-21"
output: html_document
---

This script will deal with gathering & arranging data from aphids and plants contained in excel files into structured/clean data frames (DFs) that can be used in R

#Load packages
```{r}

#set working directory to project directory

#Load packages
library(knitr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
library(readxl)

```

#Load Data
Now there is a single excel file named All_data.xls with different pieces of data in each tab, let's first explore what is there.
```{r}

#First tab contains information about the content of all the other tabs
data_key <- read_xlsx("./Data/All_data_23_07_26.xlsx", sheet="Tab_overview")
data_key

#Let's keep the first two columns to know what each tab of the excel stands for
data_key <- data_key[,1:2]

```


##DNA extraction

We'll start first with **DNA extraction**, we would like to know;

**1. Which is the median (and distribution) of DNA concentration according to the type of sample.** 
For now we'll start with the difference between leaf and epiphyte wash as we need this information to calculate the amount of plasmid to be added for spiking the samples. 

**2. If there is a correlation between the amount of leaf material weighed in the powersoil tube and the amount of DNA obtained.** [In this case, only applies to leaves as for the epiphyte wash, all of the samples had the same filter, no weighing of sample was requred.]

**3. To have an overview of which samples will be sequenced (removing those used for Glycerol stocks or those in which the amount of aphids was not enough) & to name them in a more intuitive way for sequencing and future downstream analysis

For that, we first need to **gather & arrange the data into a single df**.

```{r}

#let's see which tabs we are interested in
data_key

#ED; can tell me more information about the samples?
#DNA_C; Here is where de deNovix measurements are
#Leaf_FW; Here is where the weight of whole leaves and ground faction in powersoil tube are 

#Let's load the experimental design tab
ED <- read_xlsx("./Data/All_data_23_07_26.xlsx", sheet="ED")

#Now the DNA concentration one
DNA_C <- read_xlsx("./Data/All_data_23_07_26.xlsx", sheet="DNA_C", skip = 1) #The first row its not really the column name, so we skip it

#Finally the weight of leaves
Leaf_FW <- read_xlsx("./Data/All_data_23_07_26.xlsx", sheet="Leaf_FW") #we'll use this later 

#For now ED is not really giving anything extra, we don't need the Block information, so we'll focus on DNA_C 
str(DNA_C)
DNA_C <- DNA_C %>% select(ID, Block, Aphid_species, sample_type, fraction, sample_name,
                          experimental_batch, DNA_extraction_batch, DeNovix_ng_ul)
View(DNA_C)

#I need to have a numeric column for the concentration, but I also would like to count how many times there was the value too_low per type of sample and DNA extraction batch

DNA_C$DeNovix_ng_ul
DNA_C2 <- DNA_C %>%
  mutate(
    concentration = suppressWarnings(as.numeric(DeNovix_ng_ul)),
    too_low = DeNovix_ng_ul == "too_low",
    Glycerol = DeNovix_ng_ul == "G")

DNA_C2$Glycerol #check

#Count how many were too low per group
too_low_counts <- DNA_C2 %>%
  group_by(sample_type, DNA_extraction_batch) %>%
  summarise(
    n_too_low = sum(too_low, na.rm = TRUE),
    n_total = n(),
    .groups = "drop")

#Plot it; stacked bar
too_low_counts <- too_low_counts %>%
  mutate(n_ok = n_total - n_too_low,
    batch = factor(DNA_extraction_batch, levels = sort(unique(DNA_extraction_batch))),
    sample_type = factor(sample_type, levels = c("Blank_endo", "Blank_wash", "leaf_endo", "leaf_wash")))

ggplot(too_low_counts , aes(x = batch)) +
  geom_col(aes(y = n_ok, fill = "OK")) +
  geom_col(aes(y = n_too_low, fill = "too_low")) +
  facet_wrap(~ sample_type, nrow=1) +
  coord_cartesian(ylim = c(0, 12), )+
  scale_fill_manual(values = c("OK" = "#1b9e77", "too_low" = "#d95f02")) +
  labs(x = "DNA extraction batch", y = "Count", fill = NULL) +
  theme_bw()



```

Me estoy desesperando porque no me da el numero de samples en leaf_wash, no entiendo pq en el batch 1 tengo solo 8. Si en todos los batches fueron el mismo numero de samples, a veces una menos cuando se incluia el blank 

Necesito primero hacer un overview de cuantas samples tengo por tipo de muestra.

