---
title: "Gathering_Data"
author: "Marcela Aragon"
date: "2026-07-21"
output: html_document
---

This script will deal with gathering & arranging data from aphids and plants contained in excel files into structured/clean data frames (DFs) that can be used in R

#Load packages
```{r}

#set working directory to project directory

#Load packages
library(knitr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
library(readxl)
library(GGally) #ggpairs
library(purrr)

```

#Settings for plots
```{r}
# Setting up white background for plots
theme_set(theme_bw())

# Set default axis looks
axis_looks <- theme(axis.text.y = element_text(colour = "black", size = 10, face = "bold"))+ #text in y
  theme(axis.text.x = element_text(colour = "black", size = 12, face = "bold"))+ #text in y
  theme(axis.title.y = element_text(size = 10, face = "bold",
                                    margin = ggplot2::margin(t = 0, r = 10, b = 0, l = 0)))+
  #theme(axis.title.x = element_blank())+ #no x label
  theme(panel.grid.major = element_blank(), #no grid
        panel.grid.minor = element_blank())+ #no grid
  theme(legend.position = "top")+
  #theme(axis.ticks.x = element_blank())+  # removes x-axis ticks
  theme(
  #panel.spacing = unit(0.1, "lines"),                    # remove space between facets
  #panel.border = element_blank(),                        # remove per-facet borders
  #strip.background = element_rect(fill = "gray90", color = NA),  # keep facet titles
  panel.background = element_blank(),                    # clean background
  #panel.grid = element_blank(),                          # no grid lines
  axis.line = element_line(color = "black", linewidth = 0.3),     # axis lines restored
  axis.ticks = element_line(color = "black", linewidth = 0.3))     # axis ticks restored)  
```


#Load Data
Now there is a single excel file named All_data.xls with different pieces of data in each tab, let's first explore what is there.
```{r}

#First tab contains information about the content of all the other tabs
data_key <- read_xlsx("./Data/All_data_23_07_26.xlsx", sheet="Tab_overview")
data_key

#Let's keep the first two columns to know what each tab of the excel stands for
data_key <- data_key[,1:2]

```


##DNA extraction

We'll start first with **DNA extraction**, we would like to know;

**1. Which is the median (and distribution) of DNA concentration according to the type of sample.** 
For now we'll start with the difference between leaf and epiphyte wash as we need this information to calculate the amount of plasmid to be added for spiking the samples. 

**2. If there is a correlation between the amount of leaf material weighed in the powersoil tube and the amount of DNA obtained.** [In this case, only applies to leaves as for the epiphyte wash, all of the samples had the same filter, no weighing of sample was requred.]

**3. To have an overview of which samples will be sequenced (removing those used for Glycerol stocks or those in which the amount of aphids was not enough) & to name them in a more intuitive way for sequencing and future downstream analysis

For that, we first need to **gather & arrange the data into a single df**.

```{r}

#let's see which tabs we are interested in
data_key

#ED; can tell me more information about the samples?
#DNA_C; Here is where de deNovix measurements are
#Leaf_FW; Here is where the weight of whole leaves and ground faction in powersoil tube are 

#Let's load the experimental design tab
ED <- read_xlsx("./Data/All_data_23_07_26.xlsx", sheet="ED")

#Now the DNA concentration one
DNA_C <- read_xlsx("./Data/All_data_23_07_26.xlsx", sheet="DNA_C", skip = 1) #The first row its not really the column name, so we skip it

#Finally the weight of leaves
Leaf_FW <- read_xlsx("./Data/All_data_23_07_26.xlsx", sheet="Leaf_FW") #we'll use this later 

#For now ED is not really giving anything extra, we don't need the Block information, so we'll focus on DNA_C 
str(DNA_C)
DNA_C <- DNA_C %>% select(ID, Block, Aphid_species, sample_type, fraction, sample_name,
                          experimental_batch, DNA_extraction_batch, DeNovix_ng_ul)
View(DNA_C)

#I need to have a numeric column for the concentration, but I also would like to count how many times there was the value too_low per type of sample and DNA extraction batch

DNA_C$DeNovix_ng_ul
DNA_C2 <- DNA_C %>%
  mutate(
    concentration = suppressWarnings(as.numeric(DeNovix_ng_ul)), #make it numeric and remove those that are not numbers as NA
    too_low = DeNovix_ng_ul == "too_low", #new column
    Glycerol = DeNovix_ng_ul == "G") #new column

DNA_C2$Glycerol #check, ok
str(DNA_C2)

#Let's make factors those needed 
DNA_C2 <- DNA_C2 %>%
  mutate(across(
    c(ID,
      Block,
      Aphid_species,
      sample_type,
      fraction,
      sample_name,
      experimental_batch,
      DNA_extraction_batch,
      too_low,
      Glycerol),
    as.factor))

str(DNA_C2) #check, ok

```

Let's do a quick check of what we have 
```{r}

DNA_C2$sample_type #let's get rid of the Blank_wash and Blank_endo as it makes it confusing when calculating the amount of samples 

DNA_C2 <- DNA_C2 %>%
  mutate(
    blank = sample_type %in% c("Blank_wash", "Blank_endo"),
    sample_type = case_when(
      sample_type == "Blank_wash" ~ "leaf_wash", #change it to either leaf_wash or leaf_endo
      sample_type == "Blank_endo" ~ "leaf_endo",
      TRUE ~ as.character(sample_type)),
    sample_type = factor(sample_type),
    blank = factor(blank, levels = c(FALSE, TRUE))) #make new column to know if they are blanks

#check frequency table per factor to see how many samples are in each.
DNA_C2 %>% select(where(is.factor)) %>% map(~as.data.frame(table(.x))) #everything looks ok.

#how many samples per type of fraction?
DNA_C2 %>% count(
    experimental_batch,
    DNA_extraction_batch,
    fraction,
    sample_type) %>%  #experimental batch 5 had only 8 samples instead of 12 + four blanks, ok!
    print(n= Inf)

#How many samples in each DNA extraction batch?
DNA_C2 %>% count(
    DNA_extraction_batch,
    fraction,
    sample_type,
    Glycerol)

#Now, let's remove the glycerol samples
DNA_C3 <- DNA_C2 %>% filter(Glycerol == FALSE) %>% droplevels()

```

Too low samples 
```{r}

#How many of the samples were categorized as "too low" according to the DNA extraction batch? 
DNA_C3 %>%
  count(DNA_extraction_batch, sample_type, too_low)

#By percentage
DNA_C3 %>%
  count(DNA_extraction_batch, sample_type, too_low) %>%
  group_by(DNA_extraction_batch, sample_type) %>%
  mutate(
    total = sum(n),
    percent = round(100 * n / total, 1)
  ) %>%
  print(n = Inf)

#Stacked plot 
DNA_C3 %>%
  ggplot(aes(DNA_extraction_batch, fill = too_low)) +
  geom_bar() +
  facet_wrap(~sample_type) +
  theme_bw()

#Now if we add as well how many of those were actually blanks
ggplot(DNA_C3,
       aes(x = DNA_extraction_batch,
           fill = interaction(too_low, blank))) +
  geom_bar() +
  facet_wrap(~sample_type) +
  scale_y_continuous(breaks = seq(0, 12, by = 1)) +
  axis_looks +
  labs(fill = "Too low / Blank")


```


###DNA plot
Now I want to know what was my DNA concentration per sample type first, and then split by batch

```{r}

DNA_C3 %>%
  filter(!is.na(concentration)) %>%
  ggplot(aes(sample_type, concentration)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(
    aes(colour = blank),
    width = 0.15,
    alpha = 0.7) +
  axis_looks

#pimp it
DNA_C3 %>%
  filter(!is.na(concentration)) %>%
  ggplot(aes("", concentration)) +
  geom_boxplot(
    outlier.shape = 8,
    outlier.size = 3,
    outlier.colour = "gray"
  ) +
  geom_jitter(
    aes(colour = blank),
    width = 0.15,
    alpha = 0.7
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 3,
    fill = "gold"
  ) +
  scale_colour_manual(
    values = c("FALSE" = "black",
               "TRUE" = "red")
  ) +
  facet_wrap(~sample_type, scales = "free_y") +
  labs(x = NULL) +
  axis_looks

#Now, let's remove the outlier from leaf_wash that was very high
DNA_C4 <- DNA_C3 %>% filter(!sample_name == "1E")

#plot it again
DNA_C4 %>%
  filter(!is.na(concentration)) %>%
  ggplot(aes("", concentration)) +
  geom_boxplot(
    outlier.shape = 8,
    outlier.size = 3,
    outlier.colour = "gray"
  ) +
  geom_jitter(
    aes(colour = blank),
    width = 0.15,
    alpha = 0.7
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 3,
    fill = "gold"
  ) +
  scale_colour_manual(
    values = c("FALSE" = "black",
               "TRUE" = "red")
  ) +
  facet_wrap(~sample_type, scales = "free_y") +
  labs(x = NULL) +
  axis_looks


#check mean and median per sample_type
DNA_C4 %>%
  filter(!is.na(concentration)) %>%
  group_by(sample_type) %>%
  summarise(
    n = n(),
    mean = mean(concentration),
    median = median(concentration),
    sd = sd(concentration),
    min = min(concentration),
    max = max(concentration))

```

I want to know the concentration per type of sample and per batch to see if the first 3 were lower than 4 and 5 as I think it was.


